---
title: "Benchmarks"
author:
  - name: "Emir Turkes"
date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
link-citations: true
output:
  html_document:
    code_folding: hide
    number_sections: true
    theme: lumen
    highlight: haddock
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: false
knit: (function(inputFile, encoding) {
  rmarkdown::render(
    inputFile, encoding = encoding,
    output_file = file.path(
      "..", "results", "benchmarks.html"
    ))})
---

<style type="text/css">
body {font-size: 16px;}
h1.title {font-size: 35px;}
h1 {font-size: 24px;}
h2 {font-size: 22px;}
.toc-content {padding-left: 0px; padding-right: 0px;}
div.tocify {width: 100%;}
.tocify-subheader .tocify-item {
  font-size: 0.95em; padding-left: 25px; text-indent: 0;
}
.tocify-subheader .tocify-subheader .tocify-item {
  font-size: 0.95em; padding-left: 35px; text-indent: 0;
}
div.main-container {max-width: none; width: 100%;}
</style>

*This file is a part of [genefunnel-benchmarks](https://github.com/eturkes/genefunnel-benchmarks).*

The table of contents in the top left is clickable and can be used to quickly navigate the document.
To toggle the visibility of code, use the `CODE` toggles at the top right of chunks.
The toggle at the start of the document controls the visibility of all chunks.

```{r}
#    This file is part of genefunnel-benchmarks.
#    Copyright (C) 2024  Emir Turkes, UK DRI at UCL
#
#    This program is free software: you can redistribute it and/or modify
#    it under the terms of the GNU General Public License as published by
#    the Free Software Foundation, either version 3 of the License, or
#    (at your option) any later version.
#
#    This program is distributed in the hope that it will be useful,
#    but WITHOUT ANY WARRANTY; without even the implied warranty of
#    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
#    GNU General Public License for more details.
#
#    You should have received a copy of the GNU General Public License
#    along with this program.  If not, see <http://www.gnu.org/licenses/>.
#
#    Emir Turkes can be contacted at emir.turkes@eturkes.com

# Load required packages, suppressing startup messages.
# -----------------------------------------------------
library(conflicted)
packages <- c("GSVA", "Rcpp", "Matrix", "ComplexHeatmap", "circlize")
invisible(
  suppressPackageStartupMessages(
    lapply(packages, FUN = library, character.only = TRUE)
  )
)
sourceCpp(file.path("..", "src", "calculateScores.cpp"))
source("utils.R")
# -----------------------------------------------------

# Define global settings.
# -----------------------
knitr::opts_chunk$set(fig.width = 7, fig.height = 7, dpi = 72)
ht_opt$DIMNAME_PADDING = unit(3, units = "mm")
# -----------------------

# Useful variables.
# -----------------
cache_dir <- file.path("..", "cache")
if (!dir.exists(cache_dir)) {
  dir.create(cache_dir, recursive = TRUE)
}
# -----------------
```

# Simulated Scenarios

```{r}
vec <- c(
  rep(50, times = 10),
  rep(95, times = 5), rep(5, times = 5),
  rep(95, times = 10),
  rep(5, times = 10)
)
set.seed(1)
noise <- round(runif(40, min = -4, max = 4))

mat <- matrix(
  vec + noise,
  nrow =  10,
  ncol = 4,
  dimnames = list(
    paste("Gene", toupper(letters[1:10])),
    c("Sample 1", "Sample 2", "Sample 3", "Sample 4")
  )
)

top_anno <- HeatmapAnnotation(
  sample = anno_block(
    gp = gpar(fill = "lightgray", col = "lightgray"),
    labels = c("Sample 1", "Sample 2", "Sample 3", "Sample 4"),
    labels_gp = gpar(col = "black", fontsize = 15),
    height = unit(10, "mm")
  )
)
row_anno <- rowAnnotation(
  gene_sets = anno_block(labels = c("Gene Set X", "Gene Set Y")),
  empty = anno_empty(border = FALSE, width = unit(8, units = "mm"))
)

col_range <- c(min(mat), (min(mat) + max(mat)) / 2, max(mat))
draw(
  Heatmap(
    mat,
    name = "Gene Expression",
    col = colorRamp2(col_range, colors = c("blue", "white", "red")),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    show_column_names = FALSE,
    row_title = NULL,
    top_annotation = top_anno,
    right_annotation = row_anno,
    column_split = seq(4),
    row_split = c(rep(1, times = 5), rep(2, times = 5)),
    column_title = "Hypothetical Raw Gene Count by Sample Matrix",
    rect_gp = gpar(col = "black", lwd = 0.5),
    column_gap = unit(1, units = "mm"),
    row_names_gp = gpar(fontsize = 15),
    column_title_gp = gpar(fontsize = 18),
    row_names_side = "left",
    heatmap_legend_param = list(
      labels_gp = gpar(fontsize = 14), title_gp = gpar(fontsize = 15),
      border = "lightgrey", title_position = "topcenter",
      grid_height = unit(4, "mm"), direction = "horizontal"
    )
  ),
  heatmap_legend_side = "top"
)

seekViewport("annotation_empty_1")
loc1 = deviceLoc(x = unit(0, "npc"), y = unit(0, "npc"))
seekViewport("annotation_empty_2")
loc2 = deviceLoc(x = unit(1, "npc"), y = unit(1, "npc"))

seekViewport("global")
grid.rect(
  loc2$x * 0.96, (loc1$y + loc2$y) / 2,
  width = loc2$x - loc1$x, height = loc2$y * 1.96,
  just = c("left", "centre")
)
grid.text(
  "Gene Set Z", rot = 90,
  x = (loc1$x + loc2$x) * 0.502, y = (loc1$y + loc2$y) * 0.5
)
```

```{r, results = "hide"}
gene_sets <- list(
  `Gene Set X` = rownames(mat)[1:5], `Gene Set Y` = rownames(mat)[6:10],
  `Gene Set Z` = rownames(mat)
)

out <- vector("list", length = 6)
names(out) <- c(
  "GSVA", "ssGSEA", "ssGSEA Normalised", "PLAGE", "Z-score", "GeneFunnel"
)

tmp <- gsvaParam(mat, gene_sets)
out[[1]] <- gsva(tmp, verbose = FALSE)

tmp <- ssgseaParam(mat, gene_sets, normalize = FALSE)
out[[2]] <- gsva(tmp, verbose = FALSE)

tmp <- ssgseaParam(mat, gene_sets)
out[[3]] <- gsva(tmp, verbose = FALSE)

tmp <- zscoreParam(mat, gene_sets)
out[[4]] <- gsva(tmp, verbose = FALSE)

tmp <- plageParam(mat, gene_sets)
out[[5]] <- gsva(tmp, verbose = FALSE)

out[[6]] <- genefunnel(mat, gene_sets)
```

```{r, fig.width = 5, fig.height = 3}
top_anno <- HeatmapAnnotation(
  sample = anno_block(
    gp = gpar(fill = "lightgray", col = "lightgray"),
    labels = c("S1", "S2", "S3", "S4"),
    labels_gp = gpar(col = "black", fontsize = 15),
    height = unit(10, "mm")
  )
)

for (i in seq_along(out)) {
  col_range <- c(
    min(out[[i]]), (min(out[[i]]) + max(out[[i]])) / 2, max(out[[i]])
  )
  if (!all(is.na(col_range))) {
    ht <- draw(
      Heatmap(
        out[[i]],
        name = "Gene Set Score",
        col = colorRamp2(col_range, colors = c("blue", "white", "red")),
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        show_column_names = FALSE,
        top_annotation = top_anno,
        column_split = seq(4),
        column_title = names(out)[[i]],
        rect_gp = gpar(col = "black", lwd = 0.5),
        column_gap = unit(1, units = "mm"),
        row_names_gp = gpar(fontsize = 16),
        column_title_gp = gpar(fontsize = 18),
        row_names_side = "left",
        heatmap_legend_param = list(
          labels_gp = gpar(fontsize = 14), title_gp = gpar(fontsize = 15),
          border = "lightgrey", title_position = "topcenter",
          grid_height = unit(4, "mm"), direction = "horizontal"
        )
      ),
      heatmap_legend_side = "top",
    )
    ht
  } else {
    ht <- draw(
      Heatmap(
        out[[i]],
        name = "Gene Set Score",
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        show_column_names = FALSE,
        top_annotation = top_anno,
        column_split = seq(4),
        column_title = names(out)[[i]],
        rect_gp = gpar(col = "black", lwd = 0.5),
        column_gap = unit(1, units = "mm"),
        row_names_gp = gpar(fontsize = 16),
        column_title_gp = gpar(fontsize = 18),
        row_names_side = "left",
        heatmap_legend_param = list(
          labels_gp = gpar(fontsize = 14), title_gp = gpar(fontsize = 15),
          border = "lightgrey", title_position = "topcenter",
          grid_height = unit(4, "mm"), direction = "horizontal"
        )
      ),
      heatmap_legend_side = "top",
    )
    ht
  }
}
```

# References

This is the concluding section of the document, where we output the `sessionInfo`, and create a bibliography for works cited.

```{r}
sessionInfo()
```
